**Accuracy and precision of 16S rRNA gene sequence assignments to OTUs is more important than the stability of OTUs**

### 2015-09-29 Schloss_Cluster_PeerJ_2015.Rmd
Working on R&D section

---
title: "**The difference between precision and accuracy in assigning 16S rRNA gene sequences to operational taxonomic units**"
author: Patrick D. Schloss
date: '`r format(Sys.Date(), format="%B %d, %Y")`'
bibliography: references.bib
output:
  html_document:
    includes:
      in_header: header.tex
csl: peerj.csl
fontsize: 11pt
geometry: margin=1.0in
---

@@

## Results

***Summary of He study.*** We sought to identify the more critical analyses performed in the He study. Similar to their study, we obtained the Canadian soil dataset from Roesch et al. (REF) and processed the sequences as described in their analysis. In our opinion there were three important tests. First, was to quantify whether the clustering observed for a subset of the data represented the same clustering that was found with the full dataset. The He study found that when they used the open and closed-reference methods clusters formed using the subset data most closely resembled those of the full dataset. Among the *de novo* clustering methods they observed that the abundance-based greedy clustering (AGC) algorithm followed by single linkage (SL), distance-based greedy clustering (DGC), complete linkage (CL), and average linkage (AL) algorithms. They observed a broad range of MCC values among their AL replicates, which was not explained by the authors. Second, rarefaction curves calculated using clusters obtained using a portion of the dataset did not overlap with rarefaction curves generated using clusters generated from the full dataset. This result was originally observed in the Roesch study using the complete linkage algorithm and was reproduced in the He study where this result was more pronounced problem when using the CL, SL, and DGC algorithms relative to the other algorithms. Third, the authors attempted to describe the effects of the clustering instability on comparisons of communities. Such an analysis does not make sense for the *de novo* or open-reference approaches since the labels of the OTUs are not consistent between executions of the clustering algorithms. To overcome this, the authors selected representative sequences from OTUs and merged OTUs across different sized libraries to link OTUs. Regardless, the justification for this analysis is specious as the OTU assignments are based on the data available in the dataset when the sequences are clustered and comparing assignments in this manner are irreconcilable. We will attempt to replicate the first two observations.

***Clustering approaches vary in their stability.*** We first sought to assess calculated the MCC for for each of the clustering algorithms using 20, 40, 60, and 80% relative to the clusters formed by the algorithms using the full dataset (Figure 1A). Because a random number generator is used in some of algorithms to break ties where pairs of sequences have the same distance between them, similar to the He study, we replicated each algorithm and subsample 30 times. Across these sequencing depths, we observed that the stability of the SL and CL algorithms were highly sensitive to sampling effort relative to the AL, AGC, and DGC algorithms (Figure 1A). Our results (Figure 1B) largely confirmed those of Figure 4C of He study with one notable exception. In the He study, when comparing the OTUs formed using the AL algorithm with 60% of the data, they observed a mean MCC value of approximately 0.63 (95% confidence interval between approximately 0.15 and 0.75). In contrast, we observed a mean value of `r round(z[z$method=="an" & z$fraction==0.6, "mean"], 2)` (95% confidence interval between `r round(z[z$method=="an" & z$fraction==0.6, "lci"], 2)` and `r round(z[z$method=="an" & z$fraction==0.6, "uci"], 2)`). This result suggests that the AL algorithm was far more stabile than indicated in the He study.

***Effect of sampling depth on observed number of OTUs.*** Next, we sought to determine whether the number of OTUs observed when clustering a subset of the data was the same as that observed when rarefying the full dataset back to the number of sequences in the subset. The analyses from the He and Roesch studies both compared the number of observed OTUs from the subset to the average number of expected OTUs from the full dataset instead of to the 95% confidence interval of what would be expected from the full dataset.

***Methods vary in their accuracy.***

***Stability, accuracy, and precision affected by data quality.***

***Deep sampling of 16S rRNA genes.***
- V4 region not well represented in closed reference database

***Closed-reference clustering also has a precision problem.***

>>>>

---
title: "**Accuracy and precision of 16S rRNA gene sequence assignments to OTUs is more important than the stability of OTUs**"
author: Patrick D. Schloss
date: '`r format(Sys.Date(), format="%B %d, %Y")`'
bibliography: references.bib
output:
  html_document:
    includes:
      in_header: header.tex
csl: peerj.csl
fontsize: 11pt
geometry: margin=1.0in
---

@@

## Results and Discussion

***Summary and replication of He study.*** We sought to identify the more critical analyses performed in the He study. Similar to their study, we obtained the Canadian soil dataset from Roesch et al. (REF) and processed the sequences as described in their analysis. In our opinion there were three important tests.

First, we sought to quantify whether the clustering observed for a subset of the data represented the same clustering that was found with the full dataset. The He study found that when they used the open and closed-reference methods clusters formed using the subset data most closely resembled those of the full dataset. Among the *de novo* clustering methods they observed that the abundance-based greedy clustering (AGC) algorithm followed by single linkage (SL), distance-based greedy clustering (DGC), complete linkage (CL), and average linkage (AL) algorithms. They observed a broad range of MCC values among their AL replicates, which was not explained by the authors. We first sought to assess calculated the MCC for for each of the clustering algorithms using 20, 40, 60, and 80% relative to the clusters formed by the algorithms using the full dataset (Figure 1A). Because a random number generator is used in some of algorithms to break ties where pairs of sequences have the same distance between them, similar to the He study, we replicated each algorithm and subsample 30 times. Across these sequencing depths, we observed that the stability of the SL and CL algorithms were highly sensitive to sampling effort relative to the AL, AGC, and DGC algorithms (Figure 1A). Our results (Figure 1B) largely confirmed those of Figure 4C of He study with one notable exception. In the He study, when comparing the OTUs formed using the AL algorithm with 60% of the data, they observed a mean MCC value of approximately 0.63 (95% confidence interval between approximately 0.15 and 0.75). In contrast, we observed a mean value of `r round(z[z$method=="an" & z$fraction==0.6, "mean"], 2)` (95% confidence interval between `r round(z[z$method=="an" & z$fraction==0.6, "lci"], 2)` and `r round(z[z$method=="an" & z$fraction==0.6, "uci"], 2)`). This result suggests that the AL algorithm was far more stabile than indicated in the He study. Regardless, it supports the assertion that the clustering observed for the subset of the data does not represent the same clustering that was found with the full dataset; however, the significance of these differences is unclear.

Second, rarefaction curves calculated using clusters obtained using a portion of the dataset did not overlap with rarefaction curves generated using clusters generated from the full dataset. This result was originally observed in the Roesch study using the complete linkage algorithm and was reproduced in the He study where this result was more pronounced problem when using the CL, SL, and DGC algorithms relative to the other algorithms. The He and Roesch studies both found that the CL algorithm produced fewer OTUs in the subset than in the rarefied data. Expanding the analysis to other algorithms, the He study found that the SL algorithm produced more OTUs, the AGC produced fewer, and the other algorithms produced similar numbers of OTUs than expected when comparing the subsetted data to the rarefied data. Our results support those of these previous studies (Figure 2). It was clear that inter-method differences were generally more pronounced than the differences observed between rarefying from the full dataset and from clustering the subsetted data. The number of OTUs observed was largest using the CL algorithm, followed by the open-reference algorithm. The AL, AGC, and DGC algorithms all provided comparable numbers of OTUs. Finally, the closed-referene and SL algorithms generated the fewest number of OTUs.

Third, the authors attempted to describe the effects of the clustering instability on comparisons of communities. They used Adonis to test whether the community structure represented in subsetted communities resembled that of the full dataset. Inspection of these results indicates that although they were able to detect significant p-values, they were were of marginal biological significance. Adonis R statistics close to zero indicate the community structures from the full and subsetted datasets overlapped while values of one indicate the communities are completely different. The He study observed adonis R statistics of 0.02 (closed-reference), 0.03 (open-reference), 0.07 (CL, AGC, DGC), and 0.16 (SL and AL). Regardless of the statistical or biological significance of these results, the analysis does not make sense since the *de novo* and open-reference approaches do not consistently label the OTUs that sequences belong to when the clustering algorithms are run multiple times with different random number seeds. To overcome this, the authors selected representative sequences from OTUs and merged OTUs across different sized libraries to link OTUs. The justification for this analysis is specious as the OTU assignments are based on the data available in the dataset when the sequences are clustered and comparing assignments in this manner are irreconcilable. It is not surprising that the only analysis that did not provide a significant p-value was for the closed-reference analysis, which is the only analysis that provides consistent OTU labels. Because this analysis was so poorly designed, we did not seek to reproduce it.

It is worth noting that the entire design of the He study is artificial. First, their analysis was based on a single sample. Researchers generally have dozens or hundreds of samples that are pooled and clustered together to enable comparison across samples. Second, all of the sequence data from these datasets is pooled for a single analysis. No one would ever perform an analysis based on a subset of their data. Because of these points, the value of identifying stable OTUs is unclear.

***Methods vary in their accuracy.*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs. A method can generate highly stable OTUs, but the OTUs may be meaningless by poorly representing a specified cutoff in the assignment of sequences to those OTUs. To assess the accuracy and precision of the various methods, we made use of the pairwise distance between the unique sequences to count the number of true positives and negatives and the number of false positives and negatives for each method and sampling depth so that we could calculate the average MCC value as a measure of a method's accuracy and its variation as a measure of its precision. We made three important observations. First, each of the *de novo* methods varied in how sensitive their MCC values were to additional sequences (Figure 3A). The SL and CL algorithms were the most sensitive; however, the MCC values using 80 and 100% of the data did not vary meaningfully when using *de novo* algorithms. Second, the AL algorithm out-performed the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL. Third, with the possible exception of the CL algorithm, the MCC values for each of the only demonstrated a small amount of variation between runs of the algorithm with a different ordering of the input sequences. This indicates that although there may be variation between executions of the same algorithm, they produce OTU assignments that are equally good. Revisiting the concept of stability, we question the value of obtaining stable OTUs when the the full dataset is not optimally assigned to OTUs. Our analysis indicates that the best algorithm for assigning the Canadian soils sequences to OTUs using a 97% threshold is the AL algorithm.

***Deep sampling of 16S rRNA genes.*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering algorithms. First, the Canadian soil dataset that the He study used to analyze the stability of OTUs is one of the earliest 16S rRNA gene sequence datasets published using the 454 FLX platform. Developments in sequencing technology now permits the sequencing of millions of sequences for a study. In addition, because the original phred quality scores and flowgram data are not available, it was not possible to adequately remove sequencing errors (REFS). The large number of sequences that one would expect to remain in the dataset are likely to negatively affect the performance of all of the clustering algorithms. Second, the dataset used in the He study covered the V9 region of the 16S rRNA gene. For a variety of reasons, this region is not well represented in databases, including the reference database used by the closed and open-reference algorithms. Of the XXXXXX sequences in the default QIIME database, only XXXXXX fully cover the V9 region. Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene. With these points in mind, we compared the clustering assignment for each of these algorithms using a time series experiment that was obtained using mouse stool (REFS). The MiSeq platform was used to generate XX,XXX,XXX sequences from the V4 region of the 16S rRNA gene of XXX samples. Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02%. Although no dataset is perfect for exhaustively testing these clustering algorithms, this dataset was useful for demonstrating several points.

***Closed-reference clustering also has a precision problem.***
- V9 region not well represented in closed reference database

***Evalution of an open-source alternative to USEARCH.***
